Storm surges in the Strait of Georgia simulated with a regional model

### 2015-01-03 AO_guides/manuscriptAO.tex
reordered intro inline with Youyu's comments

%some general intro paragraph here. Need something about SoG size, geography, islands, sills, rivers, tides....
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait (Figure \ref{fig:domain}). It is on average 220 km in length and 28 km wide and its maximum depth is 420 m \citep{thomson1981oceanography}. At its southern end, the Strait of Georgia is connected to the Strait of Juan de Fuca through several passages, the largest being Haro Strait. The San Juan and Gulf Islands are scattered throughout these passages where complex bathymetry and steep topographic slopes induce vigorous tidal mixing. Many rivers feed fresh water into the Salish Sea system, the most significant source being from the Fraser River. Mixed diurnal and semi-diurnal tides are found throughout most of the Salish Sea but a semi-diurnal amphidrome is present nearby Victoria, where the tides are mainly diurnal \citep{thomson1981oceanography}. This system is home to several large coastal communities, such as Vancouver and Victoria, serves as a waterway for commercial traffic and recreational use, and supports a large and active fishing industry.

%my attempt at physical oceanography in SoG- Susan feel free to edit.
Several features related to the physical oceanography of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source of fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months \citep{thomson1981oceanography}. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation \citep{masson2004modelling}. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, set a fairly restrictive time step and impose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}. This early work pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, typically from 0.002 to 0.004, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction their M$_2$ amplitudes were too low and the M$_2$ tidal phases were too large in the Strait of Georgia.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techniques to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

@@

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk to flooding and property damage caused by storm surges, a natural hazard arising from the combination of a strong wind storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  There is also a small contribution to increased sea level due to positive sea surface height anomalies propagating northward from the equator along the eastern Pacific coast during El Nino years \citep{strub2002altimeter}.

Communities in the Strait of Georgia are particularly susceptible to storm surges in the winter months when there is an increased propensity for storms and a higher mean sea level caused by southeasterly winds and geostrophic adjustment \citep{danard2003storm}. In the summer months, the wind direction changes to northwesterly and mean sea level is lower. In British Columbia waters, the largest tidal ranges occur in June and December at the summer and winter solstices \citep{thomson1981oceanography}. The highest water levels are observed in the winter around the perigean spring tide or ``King" tide, when the sun and moon are aligned with the Earth and the moon is at its perigee. As such, storm surges in the Strait of Georgia are typically observed between November and February \citep{forseth2006adaptation}. Flooding caused by a storm surge is most likely to occur during a high spring tide \citep{abeys2011extreme}.

%storm surge modelling
Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructured grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion surface waves \citep{xu2010storm} have also been identified as important issues.  Most of these models neglect the effects of stratification, either by treating the ocean as a constant temperature and salinity or by using a two-dimensional depth-averaged set of equations, an assumption that is assumed to be valid. Further, an accurate representation of the meteorological conditions at the ocean surface is important.

Storm surge modelling in the Strait of Georgia dates back to \citet{crean1988mathematical} in a barotropic finite difference model. Their work is highlighted in reviews of storm surges in Canadian waters \citep{murty1995storm, danard2003storm}.  Compared to eastern Canada \citep{gray1984preliminary, bernier2006predicting,bobanovic2006forecasting, bernier2007tide, bernier2010tide}, relatively little has been published on storm surge modelling on the west coast of Canada in recent years. However, a storm surge forecasting program is currently operating over a domain that covers the entire British Columbia coastline \citep{StormBC}. Further, a statistical overview of past surge events in British Columbia waters suggests an increase in extreme water level events during warm El Nino periods \citep{abeys2011extreme}.

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baroclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

>>>>

%some general intro paragraph here. Need something about SoG size, geography, islands, sills, rivers, tides....
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait (Figure \ref{fig:domain}). At its southern end, the Strait of Georgia is connected to the Strait of Juan de Fuca through several passages, the largest being Haro Strait. The San Juan and Gulf Islands are scattered throughout these passages where complex bathymetry and steep topographic slopes induce vigorous tidal mixing.  Mixed diurnal and semi-diurnal tides are found throughout most of the Salish Sea but a semi-diurnal amphidrome is present nearby Victoria, where the tides are mainly diurnal \citep{thomson1981oceanography}. This system is home to several large coastal communities, such as Vancouver and Victoria, serves as a waterway for commercial traffic and recreational use, and supports a large and active fishing industry.

%my attempt at physical oceanography in SoG- Susan feel free to edit.
%Several features related to the physical oceanography of this region render it a challenge to model numerically. The outflow from the Fraser River in the late spring is the primary source of fresh water within the Strait of Georgia and leads to large vertical density gradients in the summer months \citep{thomson1981oceanography}. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation \citep{masson2004modelling}. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, set a fairly restrictive time step and impose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk to flooding and property damage caused by storm surges, a natural hazard arising from the combination of a strong wind storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  There is also a small contribution to increased sea level due to positive sea surface height anomalies propagating northward from the equator along the eastern Pacific coast during El Nino years \citep{strub2002altimeter}. Communities in the Strait of Georgia are particularly susceptible to storm surges in the winter months when there is an increased propensity for storms and a higher mean sea level caused by southeasterly winds and geostrophic adjustment \citep{danard2003storm}. In the summer months, the wind direction changes to northwesterly and mean sea level is lower. In British Columbia waters, the largest tidal ranges occur in June and December at the summer and winter solstices \citep{thomson1981oceanography}. The highest water levels are observed in the winter around the perigean spring tide or ``King" tide, when the sun and moon are aligned with the Earth and the moon is at its perigee. As such, storm surges in the Strait of Georgia are typically observed between November and February \citep{forseth2006adaptation}. Flooding caused by a storm surge is most likely to occur during a high spring tide \citep{abeys2011extreme}.

Coastal communities can prepare for and respond to storm surge hazards using predictions from storm surge models to determine if flooding conditions are probable. Like coastal ocean models, some of the complexity of a storm surge model depends on the properties of the geographic location of interest. In shallow regions, it is important to include the effects of wetting and drying which allows grid cells to alternate between land and water depending on the modelled sea surface height and depth \citep{hubbert1999storm, weisberg2006hurricane}.  Other regions require modelling over a very large domain since non-local effects from the open sea may influence surge propagation \citep{weisberg2006hurricane, lane2009verification}. Such large domains typically use nested or unstructured grids to resolve the details of complicated coastlines. Appropriate treatment of rivers \citep{flather1994storm}, accurate bathymetry and the inclusion surface waves \citep{xu2010storm} have also been identified as important issues.  Most of these models neglect the effects of stratification, either by treating the ocean as a constant temperature and salinity or by using a two-dimensional depth-averaged set of equations, an assumption that is assumed to be valid. Further, an accurate representation of the meteorological conditions at the ocean surface is important.

%storm surge modelling
Storm surge modelling in the Strait of Georgia dates back to \citet{crean1988mathematical} in a barotropic finite difference model. Their work is highlighted in reviews of storm surges in Canadian waters \citep{murty1995storm, danard2003storm}.  Compared to eastern Canada \citep{gray1984preliminary, bernier2006predicting,bobanovic2006forecasting, bernier2007tide, bernier2010tide}, relatively little has been published on storm surge modelling on the west coast of Canada in recent years. However, a storm surge forecasting program is currently operating over a domain that covers the entire British Columbia coastline \citep{StormBC}. Further, a statistical overview of past surge events in British Columbia waters suggests an increase in extreme water level events during warm El Nino periods \citep{abeys2011extreme}.

%overview oF SoG/Salish Sea models
The early modelling work of \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. %A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}.
These models pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, typically from 0.002 to 0.004, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction their M$_2$ amplitudes were too low and the M$_2$ tidal phases were too large in the Strait of Georgia.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techniques to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

@@

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. To compliment previous barotropic storm surge modelling in this region, this model employs a higher resolution and includes the effect of stratification and rivers. However, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Several simulations comparing the effects of different forcing conditions are also discussed. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.
